TRIK Studio: Technical Introduction

### 2017-02-10 2017-trikStudioEnglish/trikStudioEnglish.tex
Some bibliography changes, removed unneeded figure

Almost all aforementioned visual languages are based on control-flow computational model. Such a model is easier to understand, so it suits the purpose of education well. Nevertheless this model is not the best for programming robots since they are highly reactive by their nature: a program controlling a robot is basically a transformation of sensor signals into motor impulses. Reactive models are best expressed in data-flow languages~\cite{johnston2004advances}, where a program consists of a number of <<black boxes>>, connected with data channels. Each such <<box>> (we will call it \textit{block}) has a fixed number of inputs and outputs, and its job is to transform input data into output data (we will call \textit{tokens} data transferring through channels). For instance, each robot's sensor is represented by a single block, that simply sends tokens with sensor data to its output. Many researchers  note usability of visual data-flow languages comparing to control-flow ones~\cite{johnston2004advances}, in particular because of clear visualization of data flows.

@@

TRIK Studio is an environment that allows to program robots using diagram-based and textual languages. It 	emerged as a further evolution of QReal:Robots~\cite{terekhov2013sreda} project, developed at Software Engineering chair of Saint-Petersburg State University. The official release supports Lego Mindstorms NXT, Lego Mindstorms EV3 and TRIK robotic kits. Each one of these kits can be programmed using one of two visual languages (more simple control-flow and more complex data-flow language) or one of a number of textual ones. For Lego NXT the programmer can choose from NXT OSEK C and Russian version of C (simplified for teaching textual programming), for TRIK --- JavaScript, F\#~\cite{kirsanov2014robotics} or PascalABC.NET~\cite{doliner2014basics}, for Lego EV3 the single official language for standard firmware, EV3 virtual machine's byte code, is supported.

A program created using one of visual languages (\textit{a visual program}) could be executed in one of three modes:
\begin{itemize}
    \item debugging using a virtual simulator,
    \item debugging on a PC while sending commands to a robot via USB, Bluetooth or Wi-Fi (see Section~\ref{chapter:communications}),
    \item textual code generation mode with subsequent upload and execution of the program on the robot.
\end{itemize}

@@

\begin{itemize}
    \item The first group is for blocks implementing basic algorithmic expressions, like a start and an end of a program or a subprogram, conditions, switches, arithmetic loops, blocks for parallel execution and for working with concurrent tasks (e.g. for merging and interrupting tasks), subprogram call block and a block for textual programming.
    \item The second group combines blocks working with robot peripherals. These are actions that don't require waiting. For example, there are blocks for setting motor powers, playing sounds, accessing robot video processing capabilities, synthesizing speach by given text, controlling motor encoders and LEDs, sending messages to other robots (a part of multi-agent interaction support), working with robot file system, etc.
    \item The third group consists of blocks, that <<freeze>> the control flow. They are a timer block waiting for a given number of milliseconds (similar to \texttt{msleep} function), blocks waiting for a given value from a given sensor or from an operator game pad, and a block waiting for receiving a message from another robot. 
    \item And finally, the fourth group contains blocks that provide drawing functions for basic primitives (the drawing is being performed on a robot's screen). Such primitives are lines, rectangles, ellipses, arcs, text and images. There are parameters to change color and width of pen and brush used in drawing. There are also special blocks that control robot model marker in a 2D simulation environment. It allows robot model to leave a trace on a ground while moving, similar to the Logo turtle.
\end{itemize}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{TS_CF_Labyrinth_Diagram.png}
    \caption{The program of traversing the maze using right-hand rule. The first four blocks initialize the program: set the wall proximity variable, turn the robot motors on, after that the robot moves for one second to enter the maze. The second four blocks define the main control loop. Each ten milliseconds the robot checks using its infrared sensors if it can move forward or right (the \texttt{sensorA2 <= threshold \&\& sensorA1 <= threshold} condition). If it can move forward, the movement will be performed using proportional control over the right sensor data (the <<Regulator>> subprogram). If there are walls in front of the robot and to the right, it turn 90 degrees left (the <<Turn left>> subprogram)}.
    \label{image:TS_CF_Example}
\end{figure*}

@@

The domain-specific approach (\textit{DSM-approach})~\cite{koznov2008} was employed to create the described visual language. The editor was created using QReal DSM platform~\cite{qrealMeta,kuzenkova2013qreal}.  Language metamodel was defined using QReal's metaeditor, and a module implementing the editor was automatically generated using QReal's tools. Plugging this module into the QReal's platform core a complete visual IDE based on the given language is obtained. This IDE <<inherits>> all tools and features of the DSM platform, including modern user interface, mouse gestures recognition support for creating diagram elements~(\cite{osechkina2010gestures,osechkina2012multistroke}), copy-paste and undo/redo frameworks, zooming tools, tools for creating several types of edges on diagrams, model explorers, touch screens support and many more. According to its users TRIK Studio's user interface is much more usable and ergonomic than in any other such programming environment, and we spent only about three man-days working on it. We believe that it acknowledges the choice of QReal DSM platform as an underlying technology, but we have to note that while creating tool support for TRIK Studio numerous improvements were made to the QReal platform itself.

@@

To make transition to this <<advanced>> language simpler it maintains some kind of conceptual compatibility with the control-flow language. Blocks can also be organized in a chain with explicit passing of control, to make it possible most blocks have a special activation port that ignores all input data and just executes the block like it was in the control-flow case. So users can write programs like they used to and move on to using smaller number of blocks as they gain more experienced.

Expressive power of the language allows to use it for creating well-known complex robot control systems like Rodney Brooks's categorical architecture~\cite{brooks1986robust}, Johnathan Connell's <<colony>> architecture~\cite{connell1989colony}, Ronald Arkin's behavioral navigation~\cite{arkin1987motor}, or DAMN distributed navigation approach~\cite{rosenblatt1997damn}. Proof of this statement worth a separate paper and is not provided here. Another ideas on this matter could be found, for example, in~\cite{simpson2009toward} or~\cite{banyasad2000visual}.

@@

Making the architecture highly modular allowed to provide flexible customization of the installation process: users can explicitly select which components they are interested in and install only them (for example, if someone has only a Lego EV3 kit, he or she does not need support for Lego NXT and TRIK robots, and the target installation will not contain such files). Furthermore, such an approach is highly compatible with packet-based software install systems in Linux. The architecture also benefits from such separation since the components become very low coupled: TRIK Studio's core, for example, is very minimalistic, it does not <<know>> about all features that the IDE provides and contains only objects and interfaces common to all robotic kits and models. API of each layer is fixed and well-documented, so it allows independent community developers to create extension modules supporting new languages, robotic kits and tools.

@@

An interpreter is translating visual diagrams into a sequence of commands for a target device (see Fig.~\ref{image:interpretersTSArch}). The hierarchy of devices is build in a way that it does not matter to an interpreter if it works with a real device of a virtual one (see Fig.~\ref{image:devicesTSArch}). Each device is represented by a C++ class containing code for interaction with this device. These device classes are grouped according to robotic kits they belong to and put in plug-in modules. Device classes use communication subsystem for data interchange over physical channels (see Section~\ref{chapter:communications}) and 2D and 3D simulators API when working with virtual robot models.

@@

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.7\textwidth]{TS_Devices_Architecture.png}
    \caption{TRIK Studio's device classes hierarchy}
    \label{image:devicesTSArch}
\end{figure*}

@@

An interesting consequence of this approach is that users can change diagrams during interpretation and changes will be applied <<on the fly>>. It is a convenient behavior, for example, for selecting parameters of a PID controller in a program. Meanwhile, the execution of unchanged diagram parts is optimized: block implementation objects are cached, the same is true for abstract syntax trees and type inference information for textual programs. Thus, if the program remains unchanged while being interpreted, subsequent control flow to the same parts of the diagram will employ already created objects. Diagram validation is also performed <<on the fly>>. If some incorrect fragment is found, an error message is shown. It could be seen as a drawback since users get their error messages only when such errors are reached during interpretation, but this behavior is typical for all widely used interpreted textual languages.

@@

The simulation environment has two implementations of robot behavior models: <<ideal>> and <<realistic>>. Within the ideal model all friction forces (for floor and walls) are ignored, and having constant motor powers the robot continues to move without any acceleration or slowdown. Any, even slightest collision with a wall will stop the robot. The realistic model takes care of thrust and friction force, when crashing into a wall the robot tries to behave like the real device. Users also have an ability to add Gaussian noise to sensor data and motor impulses. Changing between these models is done using checkboxs in the simulation settings window and could be done even during the execution of a program.

@@

The constraints checking system is also used for functional testing of TRIK Studio on continuous integration server. For more information on this (and other details of TRIK Studio testing) see in~\cite{mordvinov2016testing} (in Russian).

@@

The checker can also send a visual representation of a field with a trace of a robot to a client. A trace consists of points of a robot trajectory and values reported by its sensors, and this trace can be played back by a web application described in~\cite{zakharov2016web} (in Russian). This web application also allows to actually create solutions for almost all tasks from the course right in the browser, form a correct TRIK Studio save file and run it on a server on the checker, playing back the result, so a student perceives this as "`TRIK Studio in a browser"'. This application works standalone, but is not integrated into the Stepik course yet due to technical difficulties related to the Stepik infrastructure.

>>>>

Almost all aforementioned visual languages are based on control-flow computational model. Such a model is easier to understand, so it suits the purpose of education well. Nevertheless this model is not the best for programming robots since they are highly reactive by their nature: a program controlling a robot is basically a transformation of sensor signals into motor impulses. Reactive models are best expressed in data-flow languages~\cite{johnston2004advances}, where a program consists of a number of "`black boxes"', connected with data channels. Each such "`box"' (we will call it \textit{block}) has a fixed number of inputs and outputs, and its job is to transform input data into output data (we will call \textit{tokens} data transferring through channels). For instance, each robot's sensor is represented by a single block, that simply sends tokens with sensor data to its output. Many researchers  note usability of visual data-flow languages comparing to control-flow ones~\cite{johnston2004advances}, in particular because of clear visualization of data flows.

@@

TRIK Studio is an environment that allows to program robots using diagram-based and textual languages. It 	emerged as a further evolution of QReal:Robots~\cite{terekhov2013robots} project, developed at Software Engineering chair of Saint-Petersburg State University. It supports Lego Mindstorms NXT, Lego Mindstorms EV3 and TRIK robotic kits. Each one of these kits can be programmed using one of two visual languages (more simple control-flow and more complex data-flow language) or one of a number of textual ones. For Lego NXT the programmer can choose from NXT OSEK C and Russian version of C (simplified for teaching textual programming), for TRIK --- JavaScript, F\#~\cite{kirsanov2014robotics} or PascalABC.NET~\cite{doliner2014basics}, for Lego EV3 the single official language for standard firmware, EV3 virtual machine's byte code, is supported.

A program created using one of visual languages (\textit{a visual program}) could be executed in one of three modes:
\begin{itemize}
    \item debugging using a virtual simulator,
    \item debugging on a PC while sending commands to a robot via USB, Bluetooth or Wi-Fi,
    \item textual code generation mode with subsequent upload and execution of the program on the robot.
\end{itemize}

@@

\begin{itemize}
    \item The first group is for blocks implementing basic algorithmic expressions, like a start and an end of a program or a subprogram, conditions, switches, arithmetic loops, blocks for parallel execution and for working with concurrent tasks (e.g. for merging and interrupting tasks), subprogram call block and a block for textual programming.
    \item The second group combines blocks working with robot peripherals. These are actions that don't require waiting. For example, there are blocks for setting motor powers, playing sounds, accessing robot video processing capabilities, synthesizing speach by given text, controlling motor encoders and LEDs, sending messages to other robots (a part of multi-agent interaction support), working with robot file system, etc.
    \item The third group consists of blocks, that "`freeze"' the control flow. They are a timer block waiting for a given number of milliseconds (similar to \texttt{msleep} function), blocks waiting for a given value from a given sensor or from an operator game pad, and a block waiting for receiving a message from another robot. 
    \item And finally, the fourth group contains blocks that provide drawing functions for basic primitives (the drawing is being performed on a robot's screen). Such primitives are lines, rectangles, ellipses, arcs, text and images. There are parameters to change color and width of pen and brush used in drawing. There are also special blocks that control robot model marker in a 2D simulation environment. It allows robot model to leave a trace on a ground while moving, similar to the Logo turtle.
\end{itemize}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{TS_CF_Labyrinth_Diagram.png}
    \caption{The program of traversing the maze using right-hand rule. The first four blocks initialize the program: set the wall proximity variable, turn the robot motors on, after that the robot moves for one second to enter the maze. The second four blocks define the main control loop. Each ten milliseconds the robot checks using its infrared sensors if it can move forward or right (the \texttt{sensorA2 <= threshold \&\& sensorA1 <= threshold} condition). If it can move forward, the movement will be performed using proportional control over the right sensor data (the "`Regulator"' subprogram). If there are walls in front of the robot and to the right, it turn 90 degrees left (the "`Turn left"' subprogram)}.
    \label{image:TS_CF_Example}
\end{figure*}

@@

The domain-specific approach (\textit{DSM-approach})~\cite{koznov2008} was employed to create the described visual language. The editor was created using QReal DSM platform~\cite{qrealMeta,kuzenkova2013qreal}. Language metamodel was defined using QReal's metaeditor, and a module implementing the editor was automatically generated using QReal's tools (as some other technologies like~\cite{bryksin2011ubiq}). Plugging this module into the QReal's platform core a complete visual IDE based on the given language is obtained. This IDE "`inherits"' all tools and features of the DSM platform, including modern user interface, mouse gestures recognition support for creating diagram elements~(\cite{osechkina2010gestures,osechkina2012multistroke}), copy-paste and undo/redo frameworks, zooming tools, tools for creating several types of edges on diagrams, model explorers, touch screens support and many more. According to its users TRIK Studio's user interface is much more usable and ergonomic than in any other such programming environment, and we spent only about three man-days working on it. We believe that it acknowledges the choice of QReal DSM platform as an underlying technology, but we have to note that while creating tool support for TRIK Studio numerous improvements were made to the QReal platform itself.

@@

To make transition to this "`advanced"' language simpler it maintains some kind of conceptual compatibility with the control-flow language. Blocks can also be organized in a chain with explicit passing of control, to make it possible most blocks have a special activation port that ignores all input data and just executes the block like it was in the control-flow case. So users can write programs like they used to and move on to using smaller number of blocks as they gain more experienced.

Expressive power of the language allows to use it for creating well-known complex robot control systems like Rodney Brooks's categorical architecture~\cite{brooks1986robust}, Johnathan Connell's "`colony"' architecture~\cite{connell1989colony}, Ronald Arkin's behavioral navigation~\cite{arkin1987motor}, or DAMN distributed navigation approach~\cite{rosenblatt1997damn}. Proof of this statement worth a separate paper and is not provided here. Another ideas on this matter could be found, for example, in~\cite{simpson2009toward} or~\cite{banyasad2000visual}.

@@

Making the architecture highly modular allowed to provide flexible customization of the installation process: users can explicitly select which components they are interested in and install only them (for example, if someone has only a Lego EV3 kit, he or she does not need support for Lego NXT and TRIK robots, and the target installation will not contain such files). Furthermore, such an approach is highly compatible with packet-based software install systems in Linux. The architecture also benefits from such separation since the components become very low coupled: TRIK Studio's core, for example, is very minimalistic, it does not "`know"' about all features that the IDE provides and contains only objects and interfaces common to all robotic kits and models. API of each layer is fixed and well-documented, so it allows independent community developers to create extension modules supporting new languages, robotic kits and tools.

@@

An interpreter is translating visual diagrams into a sequence of commands for a target device (see Fig.~\ref{image:interpretersTSArch}). The hierarchy of devices is build in a way that it does not matter to an interpreter if it works with a real device of a virtual one. Each device is represented by a C++ class containing code for interaction with this device. These device classes are grouped according to robotic kits they belong to and put in plug-in modules. Device classes use communication subsystem for data interchange over physical channels (see Section~\ref{chapter:communications}) and 2D and 3D simulators API when working with virtual robot models.

@@

@@

An interesting consequence of this approach is that users can change diagrams during interpretation and changes will be applied "`on the fly"'. It is a convenient behavior, for example, for selecting parameters of a PID controller in a program. Meanwhile, the execution of unchanged diagram parts is optimized: block implementation objects are cached, the same is true for abstract syntax trees and type inference information for textual programs. Thus, if the program remains unchanged while being interpreted, subsequent control flow to the same parts of the diagram will employ already created objects. Diagram validation is also performed "`on the fly"'. If some incorrect fragment is found, an error message is shown. It could be seen as a drawback since users get their error messages only when such errors are reached during interpretation, but this behavior is typical for all widely used interpreted textual languages.

@@

The simulation environment has two implementations of robot behavior models: "`ideal"' and "`realistic"'. Within the ideal model all friction forces (for floor and walls) are ignored, and having constant motor powers the robot continues to move without any acceleration or slowdown. Any, even slightest collision with a wall will stop the robot. The realistic model takes care of thrust and friction force, when crashing into a wall the robot tries to behave like the real device. Users also have an ability to add Gaussian noise to sensor data and motor impulses. Changing between these models is done using checkboxs in the simulation settings window and could be done even during the execution of a program.

@@

The constraints checking system is also used for functional testing of TRIK Studio on continuous integration server. For more information on this (and other details of TRIK Studio testing) see in~\cite{mordvinov2016testing}.

@@

The checker can also send a visual representation of a field with a trace of a robot to a client. A trace consists of points of a robot trajectory and values reported by its sensors, and this trace can be played back by a web application described in~\cite{zakharov2016web}. This web application also allows to actually create solutions for almost all tasks from the course right in the browser, form a correct TRIK Studio save file and run it on a server on the checker, playing back the result, so a student perceives this as "`TRIK Studio in a browser"'. This application works standalone, but is not integrated into the Stepik course yet due to technical difficulties related to the Stepik infrastructure.
